---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  eval = FALSE
)
```

# openalexSnapshot

<!-- badges: start -->
[![r-universe](https://openalexpro.r-universe.dev/badges/openalexSnapshot)](https://openalexpro.r-universe.dev/openalexSnapshot)
[![Codecov test coverage](https://codecov.io/gh/rkrug/openalexSnapshot/graph/badge.svg)](https://app.codecov.io/gh/rkrug/openalexSnapshot)
<!-- badges: end -->


## Installation


```{r}
install.packages(
  "openalexSnapshot",
  repos = c("https://openalexpro.r-universe.dev", "https://cloud.r-project.org")
)
```

Install the development version from GitHub:

```{r}
# install.packages("pak")
pak::pak("rkrug/openalexSnapshot")
```

## Hardware Requirements

| Resource | Minimum | Recommended |
|----------|---------|-------------|
| Disk space | 2.5 TB | 3+ TB |
| RAM | 16 GB | 32+ GB |
| CPU | 2 cores | 4+ cores |

## Quick Start

```{r}
library(openalexSnapshot)

root <- "/Volumes/openalex"

# OpenAlex now publishes the snapshot natively in Parquet, so there is no
# JSON conversion step.

# 1. Build the ID index (record lookup)
build_corpus_index(
  root_dir  = root,
  data_sets = "works",
  workers   = 4
)

# 2. Build the citation index (who cites whom). This inverts
#    referenced_works, which is what makes the forward direction possible
#    offline -- cited_by_api_url is a URL and cannot be used.
build_citation_index(root_dir = root, workers = 4)

# 3. Look up records by OpenAlex ID. `columns` matters: works carry ~51
#    columns of nested structs and extracting all of them dominates the run.
works <- lookup_by_id(
  root_dir = root,
  ids      = c("W2741809807", "W2100837269"),
  columns  = c("id", "doi", "title", "publication_year")
)

# 4. Walk the citation graph
citing <- get_citing("W2741809807", root_dir = root)   # works citing it
cited  <- get_cited("W2741809807",  root_dir = root)   # works it cites
```

## Documentation

Full documentation and articles are available at <https://openalexpro.github.io/openalexSnapshot>.

- [Working with the OpenAlex Bulk Snapshot](https://openalexpro.github.io/openalexSnapshot/articles/snapshot-workflow.html) — download, convert, index, and query the full snapshot
- [Snapshot Conversion: From JSON to Parquet](https://openalexpro.github.io/openalexSnapshot/articles/snapshot-conversion.html) — detailed function reference

## Related packages

- [openalexPro](https://openalexpro.github.io/openalexPro) — API access, tidy data frames, and advanced OpenAlex workflows
