---
title: "Working with the OpenAlex Bulk Snapshot"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
vignette: >
  %\VignetteIndexEntry{Working with the OpenAlex Bulk Snapshot}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  cache: false
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Introduction

OpenAlex provides complete data snapshots that can be downloaded and processed locally. This approach offers several advantages over API queries:

- **No rate limits**: Process millions of records without API restrictions
- **Reproducibility**: Work with a fixed point-in-time dataset
- **Speed**: Local queries are significantly faster than API calls (in many cases, especially where large result corpora are concerned)
- **Offline access**: No internet connection required after download

This vignette guides you through:

1. Setting up your environment
2. Downloading the OpenAlex snapshot
3. Converting to Parquet format
4. Building indexes for fast lookups
5. Querying your local corpus
6. Managing updates

# Prerequisites

## Hardware Requirements

Working with the full OpenAlex snapshot requires significant resources:

| Resource | Minimum | Recommended |
|----------|---------|-------------|
| Disk space | 2.5 TB | 3+ TB |
| RAM | 16 GB | 32+ GB |
| CPU | 2 cores | 4+ cores |

The snapshot download is approximately 1.2 TB. The Parquet conversion adds another 100–200 GB depending on compression settings.

## Software Requirements

- **R** (>= 4.1.2)
- **openalexSnapshot**: This package (see installation below)
- **AWS CLI**: For downloading the snapshot from S3

### Installing openalexSnapshot


```{r}
install.packages(
  "openalexSnapshot",
  repos = c("https://openalexpro.r-universe.dev", "https://cloud.r-project.org")
)
```

For the development version:

```{r}
# install.packages("pak")
pak::pak("rkrug/openalexSnapshot")
```

### Installing AWS CLI

```bash
# macOS (with Homebrew)
brew install awscli

# Ubuntu/Debian
sudo apt install awscli

# Windows (with winget)
winget install Amazon.AWSCLI
```

Verify installation:

```bash
aws --version
```

No AWS account is required — the OpenAlex bucket allows anonymous access.

# Step-by-Step Guide

## 1. Set Up Your Root Directory

Choose a location with sufficient disk space. All paths are derived from a single `root_dir`:

```
<root_dir>/
  parquet/                    # the snapshot, plus the indexes built over it
  openalex-snapshot_metadata/ # reports, logs, schema caches
```

## 2. Download the Snapshot

OpenAlex publishes the snapshot **natively in Parquet**, so there is no
conversion step: download it and you are ready to index. (Earlier releases of
this package shipped a `snapshot_to_parquet()` that converted the old
`.json.gz` distribution; it has been removed, because that path is dead.)

The snapshot is hosted on Amazon S3 and updated regularly. `aws s3 sync`
resumes interrupted downloads and only fetches new or changed files:

```bash
aws s3 sync \
  s3://openalex \
  /Volumes/external-drive/openalex/parquet \
  --no-sign-request
```

**Expected time**: 2–12 hours depending on your connection.

**Tip**: monitor progress in another terminal:

```bash
watch -n 60 'du -sh /Volumes/external-drive/openalex/parquet/*'
```

### Snapshot Structure

Each entity is a directory of Parquet files, hive-partitioned by update date:

```
parquet/
├── works/
│   └── updated_date=YYYY-MM-DD/
│       ├── part_0000.parquet
│       └── ...
├── authors/
├── institutions/
├── sources/
├── publishers/
├── funders/
└── topics/
```

## 3. A note on size

The works corpus is the bulk of it — on the order of 900 GB across ~2,000
files, roughly 500 million records. Two consequences worth knowing before you
start:

- **Column projection is the main cost lever.** Works carry ~51 columns, many
  of them deeply nested; `abstract` alone is over 10 GB. `lookup_by_id()` and
  `get_citing()` both take `columns`, and naming only what you need changes
  runtimes by an order of magnitude.
- **Point the DuckDB spill directory at a fast local disk.** Spilling onto the
  same external volume as the corpus was measured about 10x slower than onto
  an internal SSD. The builders default `temp_dir` to `tempdir()` for this
  reason; override it only if `tempdir()` is small or slow.

## 4. Build Search Indexes

For fast ID-based lookups, build an index for each dataset:

```{r}
build_corpus_index(
  root_dir  = "/Volumes/external-drive/openalex",
  data_sets = NULL,   # NULL = all datasets
  workers   = 8
)
```

This creates `<root_dir>/parquet/<dataset>_id_idx/` for each dataset — a
**partitioned directory**, not a single file, so a lookup opens only the
blocks it needs. The index maps each OpenAlex ID to its physical location
(file and row number) so `lookup_by_id()` can retrieve records without
scanning the corpus.

Two further indexes are optional but needed for the citation graph:

```{r}
# Who cites whom. This inverts `referenced_works`, and is what makes the
# forward direction possible offline at all: `cited_by_api_url` is a URL.
build_citation_index(
  root_dir = "/Volumes/external-drive/openalex",
  workers  = 8
)

# Only if you want to resolve DOIs without the API.
build_doi_index(root_dir = "/Volumes/external-drive/openalex")
```

`build_citation_index()` is the expensive one — it materialises roughly three
billion edges — and it is resumable: completed batches are skipped on a
re-run, so an interrupted build continues rather than restarting.

# Using Your Local Corpus

## Looking Up Records by ID

```{r}
library(openalexSnapshot)

# Extract specific works — output written to a temp directory
out_dir <- tempfile()
lookup_by_id(
  index_file = "/Volumes/external-drive/openalex/parquet/works_id_idx.parquet",
  ids        = c("W2741809807", "W2100837269"),
  output_dir = out_dir
)

# Read the results
library(arrow)
works <- open_dataset(out_dir) |> collect()
```

IDs can be supplied in short form (`"W2741809807"`) or long form (`"https://openalex.org/W2741809807"`); both are accepted and normalised automatically.

## Direct DuckDB Queries

For complex analytical queries, use DuckDB directly:

```{r}
library(duckdb)
library(DBI)

con <- dbConnect(duckdb())

# Register parquet files as a view
dbExecute(
  con,
  "
  CREATE VIEW works AS
  SELECT * FROM parquet_scan('/Volumes/external-drive/openalex/parquet/works/*.parquet')
  "
)

# Run analytical queries
result <- dbGetQuery(
  con,
  "
  SELECT publication_year, COUNT(*) AS n_works
  FROM works
  WHERE publication_year >= 2020
  GROUP BY publication_year
  ORDER BY publication_year
  "
)

dbDisconnect(con)
```

# Managing Updates

OpenAlex releases new snapshots regularly. The recommended update workflow:

## Checking for Updates

```bash
# Show bucket info (size, file count)
aws s3 ls s3://openalex/data/ --no-sign-request --recursive --summarize | tail -2
```

## Update Workflow

```bash
# 1. Archive the current snapshot with a datestamp
mv /Volumes/external-drive/openalex/snapshot \
   /Volumes/external-drive/openalex/snapshot-$(date +%Y-%m-%d)

# 2. Download the new snapshot
aws s3 sync \
  s3://openalex \
  /Volumes/external-drive/openalex/snapshot \
  --no-sign-request

# 3. Archive the current parquet directory
mv /Volumes/external-drive/openalex/parquet \
   /Volumes/external-drive/openalex/parquet-$(date +%Y-%m-%d)
```


# Troubleshooting

## Download Issues

**Interrupted download**: Re-run the `aws s3 sync` command. It resumes where it left off.

**Slow download**: OpenAlex uses CloudFront CDN. Speed depends on your location and internet connection. Consider running overnight.

**Disk space errors**: Ensure you have at least 400 GB free before starting.

## Conversion Issues

**Out of memory**: Reduce memory usage:

```{r}
  root_dir     = "/Volumes/external-drive/openalex",
  workers      = 1,
  profile      = "safe",
  memory_limit = 4000
)
```


# Additional Resources

- [OpenAlex Documentation](https://docs.openalex.org/)
- [OpenAlex Snapshot Documentation](https://docs.openalex.org/download-all-data/openalex-snapshot)
- [DuckDB Documentation](https://duckdb.org/docs/)
- [Apache Parquet Format](https://parquet.apache.org/)
- [openalexPro](https://openalexpro.github.io/openalexPro) — API access and advanced workflows
